Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature

### 2020-07-15 manuscript/v1_medrxiv/manuscript.Rmd
fix abstract

---
title: Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
author:
  - name: Edgar Steiger
    email: esteiger@zi.de
    affiliation: Zi
  - name: Tobias Mußgnug
    email: tmussgnug@zi.de
    affiliation: Zi
  - name: Lars Eric Kroll
    email: lkroll@zi.de
    affiliation: Zi
    corresponding: lkroll@zi.de
address:
  - code: Zi
    address: Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany
abstract: Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated their causal effects on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g. Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable causal effects of increasing temperature and of increased public mobility for essential shopping (grocery and pharmacy) as well as awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google and higher rainfall, leading to an increase in new COVID-19 cases. This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

>>>>

---
title: Causal analysis of COVID-19 observational data in German districts reveals effects of mobility, awareness, and temperature
author:
  - name: Edgar Steiger
    email: esteiger@zi.de
    affiliation: Zi
  - name: Tobias Mußgnug
    email: tmussgnug@zi.de
    affiliation: Zi
  - name: Lars Eric Kroll
    email: lkroll@zi.de
    affiliation: Zi
    corresponding: lkroll@zi.de
address:
  - code: Zi
    address: Central Research Institute of Ambulatory Health Care in Germany (Zi), Salzufer 8, D-10587 Berlin, Germany
abstract: Mobility, awareness, and weather are suspected to be causal drivers for new cases of COVID-19 infection. Correcting for possible confounders, we estimated their causal effects on reported case numbers. To this end, we used a directed acyclic graph (DAG) as a graphical representation of the hypothesized causal effects of the aforementioned determinants on new reported cases of COVID-19. Based on this, we computed valid adjustment sets of the possible confounding factors. We collected data for Germany from publicly available sources (e.g. Robert Koch Institute, Germany's National Meteorological Service, Google) for 401 German districts over the period of 15 February to 8 July 2020, and estimated total causal effects based on our DAG analysis by negative binomial regression. Our analysis revealed favorable causal effects of increasing temperature, increased public mobility for essential shopping (grocery and pharmacy), and awareness measured by COVID-19 burden, all of them reducing the outcome of newly reported COVID-19 cases. Conversely, we saw adverse effects of public mobility in retail and recreational areas, awareness measured by searches for “corona” in Google, and higher rainfall, leading to an increase in new COVID-19 cases. This comprehensive causal analysis of a variety of determinants affecting COVID-19 progression gives strong evidence for the driving forces of mobility, public awareness, and temperature, whose implications need to be taken into account for future decisions regarding pandemic management.

### 2020-07-15 manuscript/v1_medrxiv/manuscript.tex
fix abstract

  \begin{abstract}
  Mobility, awareness, and weather are suspected to be causal drivers for
  new cases of COVID-19 infection. Correcting for possible confounders, we
  estimated their causal effects on reported case numbers. To this end, we
  used a directed acyclic graph (DAG) as a graphical representation of the
  hypothesized causal effects of the aforementioned determinants on new
  reported cases of COVID-19. Based on this, we computed valid adjustment
  sets of the possible confounding factors. We collected data for Germany
  from publicly available sources (e.g.~Robert Koch Institute, Germany's
  National Meteorological Service, Google) for 401 German districts over
  the period of 15 February to 8 July 2020, and estimated total causal
  effects based on our DAG analysis by negative binomial regression. Our
  analysis revealed favorable causal effects of increasing temperature and
  of increased public mobility for essential shopping (grocery and
  pharmacy) as well as awareness measured by COVID-19 burden, all of them
  reducing the outcome of newly reported COVID-19 cases. Conversely, we
  saw adverse effects of public mobility in retail and recreational areas,
  awareness measured by searches for ``corona'' in Google and higher
  rainfall, leading to an increase in new COVID-19 cases. This
  comprehensive causal analysis of a variety of determinants affecting
  COVID-19 progression gives strong evidence for the driving forces of
  mobility, public awareness, and temperature, whose implications need to
  be taken into account for future decisions regarding pandemic
  management.
  \end{abstract}

>>>>

  \begin{abstract}
  Mobility, awareness, and weather are suspected to be causal drivers for
  new cases of COVID-19 infection. Correcting for possible confounders, we
  estimated their causal effects on reported case numbers. To this end, we
  used a directed acyclic graph (DAG) as a graphical representation of the
  hypothesized causal effects of the aforementioned determinants on new
  reported cases of COVID-19. Based on this, we computed valid adjustment
  sets of the possible confounding factors. We collected data for Germany
  from publicly available sources (e.g.~Robert Koch Institute, Germany's
  National Meteorological Service, Google) for 401 German districts over
  the period of 15 February to 8 July 2020, and estimated total causal
  effects based on our DAG analysis by negative binomial regression. Our
  analysis revealed favorable causal effects of increasing temperature,
  increased public mobility for essential shopping (grocery and pharmacy),
  and awareness measured by COVID-19 burden, all of them reducing the
  outcome of newly reported COVID-19 cases. Conversely, we saw adverse
  effects of public mobility in retail and recreational areas, awareness
  measured by searches for ``corona'' in Google, and higher rainfall,
  leading to an increase in new COVID-19 cases. This comprehensive causal
  analysis of a variety of determinants affecting COVID-19 progression
  gives strong evidence for the driving forces of mobility, public
  awareness, and temperature, whose implications need to be taken into
  account for future decisions regarding pandemic management.
  \end{abstract}
